Anxiety: a decision-theoretic perspective

### 2019-01-06 manuscript/manuscript.tex
updated/fixed cliff-walking figure.
caught up latex to gdocs

\startmain % to start the main 1-4 pages of the submission.

\section{Introduction}

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/01_field.png}}%
  }
  \caption{\textbf{Open Field Environment}}
  \par Oooh so pretty.
\end{figure}

@@

@@

@@

>>>>

\startmain

\section{Introduction}
Many psychiatric disorders involve distortions in models of the external world
and, correspondingly, aberrations of learning and decision making. As a normative
framework, decision theory allows us to describe and pose quantita­tive questions
about optimal and approximately optimal choice behavior \cite{DayanDaw2008} and
is, therefore, a critical tool for modeling, understanding, and predicting the
psychological and neurobiological mechanisms responsible for the abnormalities
observed in psychiatric behavior \cite{HuysDawDayan2015}. Many recent investigations
employing this approach have provided important insights into psychiatric symptoms,
including anhedonia in major depression \cite{Rutledge2017}, mood instability
in bipolar disorder \cite{EldarNiv2015, EldarDolanNiv2016},  habitual decision
making in disorders of compulsivity \cite{gillan2016}, and hallucinations in
schizophrenia \cite{powers2017, corlett2018}.

Anxiety disorders (generalized anxiety disorder, social anxiety disorder, panic
disorder, obsessive-compulsive disorder) represent the most common family of
psychiatric disorders (citation). Despite the differences in the content of worry,
all anxiety disorders share two core features: pessimistic inference and excessive
avoidance. Pessimistic inference describes a tendency to appraise threatening
situations as more frequent and/or more severe than they objectively are. This has
been reliably observed experimentally, as in fear conditioning wherein individuals
with anxiety exhibit exaggerated fear towards safe cues (CS-) and extinguished
threatening cues \cite{lissek2005, Duits2015}, and clinically, through self-report
(pick 3 citaitons). Excessive avoidance describes the corresponding tendency for
individuals with anxiety to consistently flee or avoid threatening situations.
Avoidance is often chosen even at the expense of potential reward (approch-avoidance
conflict), and is frequently a major impediment to recovery \cite{Arnaudova2017}.

Recent studies have attempted to provide some computational models of this behavior
\cite{Moutoussis2008, Maia2010}. For example, \cite{Maia2010} was able to show
how reinforcement learning could explain sustained avoidance in the absence of
repeated reinforcement, observed in rodents \cite{servatius2008} and something
not explained by two-factor theory \cite{Krypotos2015}. As has been discussed
elsewhere \cite{Moutoussis2017}, such accounts only explain the maintenance of
avoidance in short timescales, but do not explain why resistant to change and
why they do not go away in general. Such theories also do not explain the magnitude
of pessimistic inference, or the fear generalization. A broader account is needed.

In this article, we put forward a decision theoretic explanation of the core
symptoms of anxiety. In sequential decision environments, we show how a loss of
optimistic inference results in pessimistic inference, chronic avoidance, and
generalization of fear.

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/01_field.png}}%
  }
  \caption{\textbf{Open Field}}
  \par Oooh so pretty.
\end{figure}

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/02_cliff.png}}%
  }
  \caption{\textbf{Cliff Walking}}
  \par You better watch out YOU BETTER WATCH OUT
\end{figure}

Discuss similar observations in other tasks (Bach et al., 2014; Sheynin et al., 2015).

@@

Discuss Bach behavioral inhibition.

@@

\section{Discussion}

In this article, we showed...

Discuss previous computational approaches: This has also been previously been
investigated in previous computational modeling \cite{Huys2009}. There
the authors found that beliefs about a lack of controllability, mechanistically
modeled as corrupted priors about action-outcome contingencies, found deficient
learning in novel environments. The present results are no doubt related, but
address a different environment and a different behavior.

We discussed two mechanisms by which negative values can spread. There is a third
that we did not address here: that of the state itself. Here we assumed that states
were fully observable; in other words, there was no uncertainty about the state
an agent was currently occupying. In partially-observable Markov decision problems,
the state itself must be inferred from external and internal variables. This is
yet another means by which bad value can spread if undue likelihood is granted
to undesirable or aversive states. This ideas has previously been proposed to
explain some symptoms in anxiety \cite{Paulus2012}, and not without good reason.
Anxiety disorders involve negatively biased interpretation of ambiguous percepts
\cite{Hartley2012}, which may result in aberrant planning and increased avoidance
for similar reasons to those described above. This is an area for future research.

In this article, we have explored a computational principle of approach and
avoidance learning in sequential decision environments. We have not discussed,
however, the particular psychological or neurobiological mechanisms by which this
sort of learning may take place. This issue has been discussed in detail elsewhere
\cite{Bishop2018}, but we briefly discuss a few possibilities here. Mechanically,
we understand that complex decision problems are solved by mental processes that
exist on a spectrum between model-based and model-free (Daw et al., 2005). The
difference between these two families of methods hinges on the reliance of an
internal model of the world used to make predictions about expected long-run
outcomes of particular choices.

For model-based decision making, one possibility is biased planning. In large
state spaces, online sequential decision making suffers from the curve of
dimensionality: in planning a series of actions to take, comparing between
multiple options rapidly becomes computationally intractable when the number of
options and depth of search becomes even moderately large. Thus, heuristics for
reducing the size of this search can be useful. Biased pruning of search paths
(such as immediately discarding paths that require some minor loss in pursuit of
larger gains) may result in maladaptive decision making (Huys et al., 2012).
Alternately, if planning relies on internal sampling from previously experienced
episodes in order to make predictions about future value, then biased sampling
 may similarly result in maladaptive decision making. Recent work has shown how
 finite sampling during planning can result in risk aversion and the overrepresentation
 of low probability but strongly negative outcomes (Lieder et al., 2018). Such a
 mechanism is prima facie in line with results showing the availability in memory
 of negative outcomes being higher for patients with anxiety (Borkovec et al., 1999;
 Miranda and Mennin, 2007).

Within the realm of model-free decision making, several studies have found evidence
suggesting increased learning rates for negative reward prediction errors in
individuals with elevated levels of anxiety \cite{Aylward, Huang2017, Harle2017
garrett2018}. Though an asymmetry in sensitivity to
positive and negative reward prediction errors bears superficial similarity, for
the reasons described above merely attending more to negative information does not
necessarily predict increased avoidance insofar that avoidance can successfully
isolate negative outcomes. One possibility, and topic of future research, is how
prior beliefs about the reward statistics of the current environment dictate
learning rates. A second possibility is prioritized replay (Russek et al., 2017;
Mattar and Daw, 2018). Offline replay of previous experiences (through hippocampal
mechanisms) are known to facilitate learning. It has been proposed that biased
replay could bias value estimates (Gagne et al., 2018). This is another exciting
area for future research.

\section{References}

\bibliographystyle{authordate1}
\bibliography{citations}
